%%!TEX root = the-volume.tex

\chapter{Limit spaces}\label{chap:lim}\label{chap:stability}

In this lecture we show that lower curvature bound in the sense of Alexandrov survives under Gromov--Hausdorff limit and prove the Gromov selection theorem.
This theorem is the main source of applications of Alexandrov geometry,
as an illustration we prove the homotopy stability theorem (\ref{thm:h-stability}) and deduce the homotopy finiteness theorem (\ref{thm:h-finiteness}) from it.


\section{Survival of curvature bounds}

\begin{thm}{Theorem}\label{thm:CBB-closed}
Let $\spc{X}_n\z\to \spc{X}_\infty$ be a convergence in the sense of Gromov--Hausdorff.
Suppose that each for each $n$, the space $\spc{X}_n$ has curvature $\ge\kappa$ in the sense of Alexandrov.
Then the same holds for~$\spc{X}_\infty$.
\end{thm}

\parit{Proof}.
Choose a quadruple of points $p_\infty, x_\infty,y_\infty,z_\infty\in \spc{X}_\infty$.

By the definition of Gromov--Hausdorff convergence, we can choose points $p_n$,  $x_n$, $y_n$, $z_n\in \spc{X}_n$ for each $n$
that converge to $p_\infty$, $x_\infty$, $y_\infty$, $z_\infty\in \spc{X}_\infty$, respectively.
In particular, each of the 6 distances between pairs of $p_n$, $x_n$, $y_n$, $z_n$ converge to the distance between the corresponding pairs of $p_\infty, x_\infty,y_\infty,z_\infty$.

Since $\MM^2(\kappa)$-comparison holds for $p_n$, $x_n$, $y_n$, $z_n\z\in \spc{X}_n$,
passing to the limit, we get the $\MM^2(\kappa)$-comparison for $p_\infty$,  $x_\infty$, $y_\infty$, $z_\infty$.
\qeds

\begin{thm}{Exercise}\label{ex:dim-lim}
Suppose that a sequence $\spc{A}_1,\spc{A}_2,\dots$ of $\Alex\kappa$ spaces that converges to $\spc{A}_\infty$ in the sense of Gromov--Hausdorff.
Show that $\spc{A}_\infty$ is $\Alex\kappa$ and
\[\dim \spc{A}_\infty\le \liminf_{n\to\infty} \dim \spc{A}_n.\]
\end{thm}

\section{Gromov's selection theorem}

\begin{thm}{Theorem}\label{thm:gromov-compactness}
Let $D,\kappa\in\RR$, and $m$ be a positive integer. 
Then any sequence of $m$-dimensional $\Alex\kappa$ spaces with diameters at most $D$
has a converging subsequence in the sense of Gromov--Hausdorff.
\end{thm}

Let $X$ be a subset of a metric space $\spc{W}$.
Recall that a set $Z\subset \spc{W}$ is called an \index{$\eps$-net}\emph{$\eps$-net} of $X$ if for any point $x\in X$, there is a point $z\in Z$ such that $\dist{x}{z}{}<\eps$.

We will use the following characterization of compact sets.

\begin{thm}{Exercise}\label{ex:net}
A closed subset $X$ of a complete metric space.

\begin{subthm}{ex:net:finite}
Show that $X$ is compact if and only if it admits a finite $\eps$-net for any $\eps>0$.
\end{subthm}

\begin{subthm}{ex:net:compact}
Show that $X$ is compact if and only if it admits a compact $\eps$-net for any $\eps>0$.
\end{subthm}

\end{thm}

Recall that $\pack_\eps\spc{X}$ is the exact upper bound on the number of points $x_1,\z\dots,x_n\in \spc{X}$ such that $\dist{x_i}{x_j}{}\ge\eps$ if $i\ne j$.

If $n=\pack_\eps\spc{X}<\infty$, then
the collection of points $x_1,\dots,x_n$ is called a \index{maximal packing}\emph{maximal $\eps$-packing}.

\begin{thm}{Exercise}\label{ex:pack-net}
Show that any maximal $\eps$-packing $x_1,\dots,x_n$ is an $\eps$-net.
Conclude that a complete metric space $\spc{X}$ is compact if and only if $\pack_\eps\spc{X}\z<\infty$ for any $\eps>0$.
\end{thm}


\parit{Proof of \ref{thm:gromov-compactness}.}
Denote by $\bm{K}$ the set of isometry classes of $\Alex0$ spaces with dimension $\le m$ and diameter $\le D$.
By \ref{ex:dim-lim}, $\bm{K}$ is a closed subset of $\GH$.

Choose a space $\spc{A}\in \bm{K}$;
suppose $x_1,\dots,x_n\in \spc{A}$ is a collection of points such that $\dist{x_i}{x_j}{}\ge \eps$ for all $i\ne j$.
Note that the balls $B_i=\oBall(x_i,\tfrac\eps2)$ do not overlap.

By \ref{thm:right-inverse}, $\vol \spc{A}>0$.
By Bishop--Gromov inequality, $\vol \spc{A}<\infty$,
and if $\eps<D$, then 
\[\vol B_i\ge (\tfrac\eps{2\cdot D})^m\cdot\vol \spc{A}\]
for any $i$.
It follows that $n\le (\tfrac{2\cdot D}\eps)^m$;
that is, 
\[\pack_\eps\spc{A}\le  N(\eps)\df(\tfrac{2\cdot D}\eps)^m\]
for all small $\eps>0$.

Choose a maximal $\eps$-packing $x_1,\z\dots,x_n\in \spc{A}$.
By \ref{ex:pack-net}, $\spc{F}_\eps\z\df\{x_1,\z\dots,x_n\}$ is an $\eps$-net of $\spc{A}$.
Observe that $\dist{\spc{F}_\eps}{\spc{A}}{\GH}\le \eps$.
Further, note that the set $\bm{F}_\eps$ of finite metric spaces with diameter $\le D$ and at most $N(\eps)$ points forms a compact subset in $\GH$.

Summarizing, for any $\eps>0$ we can find a compact $\eps$-net $\bm{F}_\eps\subset \GH$ of $\bm{K}$.
Since $\GH$ is complete (\ref{prop:complete}), it remains to apply \ref{ex:net:compact}.

Rescaling reduces $\Alex\kappa$ case to the $\Alex{-1}$ case.
The latter can be proved in the same way, using \ref{ex:BG} instead of \ref{inq:BG}.
\qeds

\begin{thm}{Exercise}\label{ex:pack-vol}

\begin{subthm}{ex:pack-vol:pack}
Let $\spc{A}$ be an $m$-dimensional $\Alex0$ space with diameter $\le D$.
Suppose $\vol\spc{A}\ge v_0>0$.
Show that 
\[\pack_\eps\spc{A}\ge \frac\Const{\eps^m}\]
for some constant $\Const=\Const(m,D,v_0)>0$.
\end{subthm}


\begin{subthm}{ex:pack-vol:dim}
Conclude that if $\spc{A}_n$ is a sequence of $m$-dimensional $\Alex0$ spaces with diameter $\le D$, and volume $\ge v_0$, then its Gromov--Hausdorff limit $\spc{A}_\infty$ (if it exists) has dimension~$m$.
\end{subthm}
\end{thm}

\begin{thm}{Exercise}\label{ex:diam-compact:GH}
Let $(\spc{A}_1,p_1),(\spc{A}_2,p_2),\dots$ be a sequence of $m$-dimensional $\Alex\kappa$ spaces with marked points.
Show that it contains a subsequence pointed-converging in the sense of Gromov--Hausdorff.

\end{thm}
